---
title: "2_2.Data_analysis"
author: "Jaeseok Hwang"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = FALSE, echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 6
)

library(here)
library(data.table)
library(tidyverse)
library(grf)
library(fixest)
library(moments)
library(pracma)
library(ggplot2)
library(huxtable)
library(mgcv)
library(plm)
library(gamlss)
library(gamlss.dist)
library(purrr)
library(plotly)
library(stats4)
library(MASS)
library(scales)


```

## Step: Simulate 1,000 Fields with 200 Subplots Each

```{r simulate-1000-fields, message=FALSE, warning=FALSE}

# combined OFPE data from IL (2016 - 2023)
dat_comb <- readRDS(here("Data", "Processed", "Analysis_ready", "df_il.rds"))

# Select variables for regressions
dat_comb <- dat_comb %>% 
        data.table() %>%
   .[, .(yield, n, sand, clay, silt, water_storage, elev,
      slope, aspect, prcp_t, gdd_t, edd_t, ffy_id)] %>%
        na.omit()

range_check <- dat_comb[, .(n_range = max(n) - min(n)), by = ffy_id]
valid_ids <- range_check[n_range >= 45]$ffy_id
dat_tab <- dat_comb[ffy_id %in% valid_ids]




quantile(range_check$n_range)
# Step 1: Fit field-specific GAM with nitrogen + field-level covariates (NO weather)
fit_gam_by_field <- function(df) {
  tryCatch({
    gam(yield ~ s(n, k = 4) + sand + silt + clay + water_storage +
        elev + slope + aspect, data = df, method = "REML")
  }, error = function(e) NULL)
}

gam_list <- dat_tab[, .(model = list(fit_gam_by_field(.SD))), by = .(ffy_id, prcp_t)]
gam_list <- gam_list[!sapply(gam_list$model, is.null)]


# --- STEP 2: Fit field-specific yield~N curves and stratify by observed weather (prcp_t) ---

# Define GAM fitting function for each field-year
fit_gam_by_field <- function(df) {
  tryCatch({
    gam(yield ~ s(n, k = 4) + sand + silt + clay + water_storage +
        elev + slope + aspect, data = df, method = "REML")
  }, error = function(e) NULL)
}

# Fit a yield~N curve for each field-year (ffy_id) and preserve precipitation value
# NOTE: prcp_t is implicitly embedded in the yield response because it's the actual weather
# under which the field trial was conducted. Hence, each fitted curve reflects N responsiveness
# conditioned on that specific prcp_t. (See Lobell et al. 2014; Hatfield & Prueger 2015)

gam_list <- dat_tab[, .(model = list(fit_gam_by_field(.SD))), by = .(ffy_id, prcp_t)]
gam_list <- gam_list[!sapply(gam_list$model, is.null)]  # remove failed fits

# Create prediction grid to represent fitted yield~N curves for each field
n_grid <- seq(60, 220, by = 5)

# Calculate global medians for covariates
cov_meds <- dat_tab[, .(
  sand = median(sand, na.rm = TRUE),
  silt = median(silt, na.rm = TRUE),
  clay = median(clay, na.rm = TRUE),
  water_storage = median(water_storage, na.rm = TRUE),
  elev = median(elev, na.rm = TRUE),
  slope = median(slope, na.rm = TRUE),
  aspect = median(aspect, na.rm = TRUE)
)]

# Create prediction function with fixed covariates
get_fitted_curve <- function(model, cov_row) {
  data.table(
    n = n_grid,
    y = predict(model, newdata = data.frame(
      n = n_grid,
      sand = cov_row$sand,
      silt = cov_row$silt,
      clay = cov_row$clay,
      water_storage = cov_row$water_storage,
      elev = cov_row$elev,
      slope = cov_row$slope,
      aspect = cov_row$aspect
    ))
  )
}

cov_row <- as.list(sample_covariates(1)[1, ])

# Store each fitted curve along with its associated precipitation

curves <- gam_list[, .(
  curve = list(get_fitted_curve(model[[1]], cov_row))
), by = .(ffy_id, prcp_t)]

# -----------------------------
# 📌 AGRONOMY RATIONALE (Comment)
# -----------------------------

# Why this works:
# In crop systems, the shape of the yield response to nitrogen is not fixed but
# it varies substantially with precipitation and temperature due to:
# - drought stress limiting uptake at low prcp_t
# - waterlogging or leaching at high prcp_t
# - microbial nitrogen losses via denitrification (Hatfield & Prueger, 2015)

# Rather than impose a specific functional form (e.g., yield ~ N + prcp_t + N:prcp_t),
# this method learns f_i(N) flexibly from each field’s actual trial year and weather.

# By retaining `prcp_t` as an index, we can later sample yield-N response curves
# from a diversity of weather conditions (Lobell et al., 2014; Archontoulis & Licht, 2017).

# Citations:
# - Lobell, D. B., Schlenker, W., & Costa-Roberts, J. (2014). Climate Trends and Global Crop Production Since 1980. *Science*, 333(6042), 616–620.
# - Hatfield, J. L., & Prueger, J. H. (2015). Temperature extremes: Effect on plant growth and development. *Weather and Climate Extremes*, 10, 4–10.
# - Archontoulis, S. V., & Licht, M. A. (2017). How water and temperature drive crop development. *Iowa State University Extension Agronomy Technical Note*, AG 0003.

# Step 3: Fit multivariate normal for soil/topo covariates
covariate_vars <- c("sand", "silt", "clay", "water_storage", "elev", "slope", "aspect")
cov_dt <- dat_tab[, ..covariate_vars]
mu_cov <- colMeans(cov_dt)
Sigma_cov <- cov(cov_dt)

sample_covariates <- function(n_fields) {
  var_names <- c("sand", "silt", "clay", "water_storage", "elev", "slope", "aspect")
  samples <- MASS::mvrnorm(n = n_fields, mu = mu_cov, Sigma = Sigma_cov)
  data.table::setnames(as.data.table(samples), var_names)
}

#Step 4: Simulate synthetic fields using sampled weather + yield-N curves
n_fields <- 1000
n_subplots <- 200
set.seed(123)

cov_samples <- sample_covariates(n_fields)
field_fixed_effects <- rnorm(n_fields, 0, 10)

synthetic_fields <- vector("list", n_fields)

for (i in seq_len(n_fields)) {
  # Sample covariates and weather
  cov_row <- cov_samples[i]
  prcp_i <- runif(1, 400, 800)

  # Get fitted curve with covariates
  sampled_model <- gam_list[sample(.N, 1)]$model[[1]]
  sampled_curve <- get_fitted_curve(sampled_model, cov_row)

  n_values <- runif(n_subplots, 60, 220)
  y_true <- approx(x = sampled_curve$n, y = sampled_curve$y, xout = n_values)$y

  y_obs <- y_true + field_fixed_effects[i] + rnorm(n_subplots, 0, 5)

  synthetic_fields[[i]] <- data.table(
    field_id = i,
    n = n_values,
    prcp_t = prcp_i,  # Add weather info
    yield_true = y_true,
    yield = y_obs,
    fe = field_fixed_effects[i],
    sand = cov_row$sand,
    silt = cov_row$silt,
    clay = cov_row$clay,
    water_storage = cov_row$water_storage,
    elev = cov_row$elev,
    slope = cov_row$slope,
    aspect = cov_row$aspect
  )
}

subplot_dt <- rbindlist(synthetic_fields)
saveRDS(subplot_dt, here("Data", "Processed", "simul_field_1000_weather.rds"))



```

# Check the yield N response by field at 1,000 simulated fields

```{r plot simulate-1000-fields, message=FALSE, warning=FALSE}

# For color grouping: bin precipitation into 5 levels

subplot_dt[, prcp_group := cut(prcp_t, breaks = quantile(prcp_t, probs = seq(0, 1, length.out = 6)),
                               include.lowest = TRUE, labels = paste0("P", 1:5))]

# Assign each prcp_group a base color
prcp_colors <- c("P1" = "#1f77b4", "P2" = "#2ca02c", "P3" = "#ff7f0e", "P4" = "#d62728", "P5" = "#9467bd")

# Sample 500 fields for performance (optional)
# subplot_sample <- subplot_dt[sample(unique(field_id), 500), on = .(field_id)]

# Plot all 1,000 fields
ggplot(subplot_dt, aes(x = n, y = yield_true, group = field_id, color = prcp_group)) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 4), se = FALSE, size = 0.6, alpha = 0.8) +
  scale_color_manual(values = prcp_colors, name = "Precipitation Level") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Smoothed Yield–Nitrogen Response Curves by Field",
    subtitle = "Grouped by Precipitation Level (prcp_t)",
    x = "Nitrogen Rate (lbs/ac)",
    y = "Yield (bu/ac)"
  )

```

### 💡 Notes:
- The resulting `subplot_dt` has 1,000 fields × 200 subplots = **200,000 observations**
- Each field is created by:
  - Sampling a yield-N function from real fitted GAMs
  - Sampling realistic field covariates and a latent productivity term
  - Randomizing nitrogen rates across 200 subplots
- This data can now be used to **test the bias and variance** of EONR estimates under subplot- vs. field-level modeling

