---
title: "2_2.Data_analysis"
author: "Jaeseok Hwang"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
  word_document: defaulta
---

## Knitr option

```{r, cache = F, echo = F, results = "hide"}
#####

library(knitr)
knitr::opts_chunk$set(
  cache = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 6
)
```

## Load packages

```{r packages}
library(here)
library(data.table)
library(tidyverse)
library(dplyr)
library(grf)
library(fixest)
library(moments)
library(pracma)
library(ggplot2)
library(huxtable)
library(mgcv)  # For GAM
library(plm)
library(gamlss)
library(gamlss.dist)
library(purrr)
library(plotly)
library(pracma)     # for numerical integration
library(stats4)     # for MLE if needed
```


# Data analysis Outline

# 1. Estimate 1st,2nd,3rd moment

## Regress yield, yield^2 , yield^3 on nitrogen + soil, topography + weather
### by OLS and GAM
## ( field level VS subplot level)

## Predict mu1,mu2,mu3 (moments) with the regression model 
## by setting soil, topography, gdd and edd value at the median level

# 2. Estimate distribution of yield responses to nitrogen by precipitation level 
#   by Maximum_Entropy approach 

# 3. Plot(visualize) dist of yield responses to nitrogen by precipitation level

# 4. Estimate EONR under the prob density of yield by N



## Load and normalize data

```{r load-data}

dat_com <- readRDS(here("Data", "Processed", "Analysis_ready", "df_il.rds")) %>%
  data.table() %>%
  .[, .(yield, n, sand, clay, silt, water_storage, elev,
        slope, aspect, prcp_t, gdd_t, edd_t, ffy_id)] %>%
  na.omit()

yield_min <- min(dat_com$yield, na.rm = TRUE)
yield_max <- max(dat_com$yield, na.rm = TRUE)
normalize_yield <- function(y) (y - yield_min) / (yield_max - yield_min)
dat_com[, yield_scaled := normalize_yield(yield)]
dat_com[, yield_m1 := yield_scaled]
dat_com[, yield_m2 := yield_scaled^2]
dat_com[, yield_m3 := yield_scaled^3]

```

## Create field-level data from subplot data
# set field specific values at median level and get unique weather values 
```{r field-level-data}

field_dt <- dat_com[, .(
  yield_m1 = median(yield_scaled),
  yield_m2 = median(yield_scaled^2),
  yield_m3 = median(yield_scaled^3),
  n = median(n),
  sand = median(sand), clay = median(clay), silt = median(silt),
  water_storage = median(water_storage), elev = median(elev),
  slope = median(slope), aspect = median(aspect),
  prcp_t = unique(prcp_t), gdd_t = unique(gdd_t), edd_t = unique(edd_t)
), by = ffy_id] %>% na.omit()

```


# 1. Estimate First, Second, Third Raw Moments

```{r subplot-moment}

# By OLS

mod_m1 <- lm(yield_m1 ~ n + sand + clay + silt + water_storage +
                  elev + slope + aspect + prcp_t + gdd_t + edd_t , data = dat_com)

mod_m2 <- lm(yield_m2 ~ n + sand + clay + silt + water_storage +
                  elev + slope + aspect + prcp_t  + gdd_t + edd_t, data = dat_com)

mod_m3 <- lm(yield_m3 ~ n + sand + clay + silt + water_storage +
                  elev + slope + aspect + prcp_t + gdd_t + edd_t, data = dat_com)

# By GAM

gam_mu1 <- gam(yield_m1 ~ s(n, k = 4) + s(prcp_t, k = 4) + 
  sand + clay + silt + water_storage + elev + slope + aspect + gdd_t + edd_t,
  data = dat_com, method = "REML")

gam_mu2 <- gam(I(yield_m2) ~ s(n, k = 4) + s(prcp_t, k = 4) + 
  sand + clay + silt + water_storage + elev + slope + aspect + gdd_t + edd_t,
  data = dat_com, method = "REML")

gam_mu3 <- gam(I(yield_m3) ~ s(n, k = 4) + s(prcp_t, k = 4) +
  sand + clay + silt + water_storage + elev + slope + aspect + gdd_t + edd_t,
  data = dat_com, method = "REML")

```

## Field-level OLS moment regressions


```{r field-moment-ols}

mod_fm1 <- lm(yield_m1 ~ n + sand + clay + silt + water_storage +
                elev + slope + aspect + prcp_t + gdd_t + edd_t, data = field_dt)
mod_fm2 <- lm(yield_m2 ~ n + sand + clay + silt + water_storage +
                elev + slope + aspect + prcp_t + gdd_t + edd_t, data = field_dt)
mod_fm3 <- lm(yield_m3 ~ n + sand + clay + silt + water_storage +
                elev + slope + aspect + prcp_t + gdd_t + edd_t, data = field_dt)

                # By GAM

gam_fm1 <- gam(yield_m1 ~ s(n, k = 4) + s(prcp_t, k = 4) + 
  sand + clay + silt + water_storage + elev + slope + aspect + gdd_t + edd_t,
  data = field_dt, method = "REML")

gam_fm2 <- gam(I(yield_m2) ~ s(n, k = 4) + s(prcp_t, k = 4) + 
  sand + clay + silt + water_storage + elev + slope + aspect + gdd_t + edd_t,
  data = field_dt, method = "REML")

gam_fm3 <- gam(I(yield_m3) ~ s(n, k = 4) + s(prcp_t, k = 4) + 
  sand + clay + silt + water_storage + elev + slope + aspect + gdd_t + edd_t,
  data = field_dt, method = "REML")
```



# 2. Create Prediction Grid

```{r subplot prediction}

# Prediction grid (identical sequence for both subplot and field_dt)
n_seq <- quantile(dat_com$n, probs = seq(0.1, 0.9, length.out = 5), na.rm = TRUE)
prcp_seq <- quantile(dat_com$prcp_t, probs = seq(0.1, 0.9, length.out = 5), na.rm = TRUE)



# Median values for other covariates
cov_meds <- dat_com[, .(sand = median(sand), clay = median(clay), silt = median(silt),
             water_storage = median(water_storage), elev = median(elev),
            slope = median(slope), aspect = median(aspect),
              gdd_t = median(gdd_t),edd_t = median(edd_t))]


grid <- CJ(n = n_seq, prcp_t = prcp_seq)
grid <- cbind(grid, cov_meds[rep(1, nrow(grid))])

# Predict Raw Moments by OLS

grid[, mu1 := predict(mod_m1, newdata = grid)]
grid[, mu2 := predict(mod_m2, newdata = grid)]
grid[, mu3 := predict(mod_m3, newdata = grid)]

# Predict Raw moments by GAM
grid[, mu1_gam := predict(gam_mu1, newdata = grid)]
grid[, mu2_gam := predict(gam_mu2, newdata = grid)]
grid[, mu3_gam := predict(gam_mu3, newdata = grid)]


grid[, var_check := mu2 - mu1^2]
summary(grid$var_check)
grid[var_check <= 0]


```

```{r field-prediction-grid}

field_cov_meds <- field_dt[, .(
  sand = median(sand), clay = median(clay), silt = median(silt),
  water_storage = median(water_storage), elev = median(elev),
  slope = median(slope), aspect = median(aspect),
  gdd_t = median(gdd_t), edd_t = median(edd_t))]

field_grid <- CJ(n = n_seq, prcp_t = prcp_seq)
field_grid <- cbind(field_grid, field_cov_meds[rep(1, nrow(field_grid))])

field_grid[, mu1 := predict(mod_fm1, newdata = field_grid)]
field_grid[, mu2 := predict(mod_fm2, newdata = field_grid)]
field_grid[, mu3 := predict(mod_fm3, newdata = field_grid)]

field_grid[, mu2 := pmax(mu2, mu1^2 + 1e-4)]
field_grid[, mu3 := pmin(pmax(mu3, mu2 * mu1 + 1e-4), mu1)]

# Predict field-level GAM moments
field_grid[, mu1_gam := predict(gam_fm1, newdata = field_grid)]
field_grid[, mu2_gam := predict(gam_fm2, newdata = field_grid)]
field_grid[, mu3_gam := predict(gam_fm3, newdata = field_grid)]

# Regularize moments to prevent invalid inputs
field_grid[, mu2_gam := pmax(mu2_gam, mu1_gam^2 + 1e-4)]
field_grid[, mu3_gam := pmin(pmax(mu3_gam, mu2_gam * mu1_gam + 1e-4), mu1_gam)]


```

## 3. Maximum Entropy Estimation Based on OLS and GAM Moments

```{r mbme-subplot-ols-vs-gam}

# Define MBME estimator function
estimate_mbme_density <- function(mu1, mu2, mu3, support = c(0, 1)) {
  if (any(is.na(c(mu1, mu2, mu3))) || (mu2 - mu1^2 <= 0)) {
    warning("Invalid or inconsistent moment inputs.")
    return(NULL)
  }

  safe_exp <- function(x) {
    x <- ifelse(x > 700, 700, x)  # avoid overflow in exp()
    exp(x)
  }

  # Objective: match estimated vs. target raw moments over [0, 1]
  f_obj <- function(lambda) {
    Z <- tryCatch(
      integrate(function(y) safe_exp(lambda[1]*y + lambda[2]*y^2 + lambda[3]*y^3), 0, 1)$value,
      error = function(e) Inf
    )
    m1 <- tryCatch(
      integrate(function(y) y * safe_exp(lambda[1]*y + lambda[2]*y^2 + lambda[3]*y^3), 0, 1)$value / Z,
      error = function(e) Inf
    )
    m2 <- tryCatch(
      integrate(function(y) y^2 * safe_exp(lambda[1]*y + lambda[2]*y^2 + lambda[3]*y^3), 0, 1)$value / Z,
      error = function(e) Inf
    )
    m3 <- tryCatch(
      integrate(function(y) y^3 * safe_exp(lambda[1]*y + lambda[2]*y^2 + lambda[3]*y^3), 0, 1)$value / Z,
      error = function(e) Inf
    )
    sum((c(m1, m2, m3) - c(mu1, mu2, mu3))^2)
  }

  # First try with BFGS, then fallback to box-constrained L-BFGS-B if needed
  res <- tryCatch({
    optim(par = c(0, 0, 0), fn = f_obj, method = "BFGS")
  }, error = function(e) NULL)

  if (is.null(res) || res$convergence != 0) {
    res <- tryCatch({
      optim(par = c(0, 0, 0), fn = f_obj, method = "L-BFGS-B",
            lower = rep(-50, 3), upper = rep(50, 3))
    }, error = function(e) NULL)
  }

  if (is.null(res) || res$convergence != 0) {
    warning("MBME optimization failed — returning NULL.")
    return(NULL)
  }

  lambda <- res$par

  # Return MBME density function on [support]
  function(y) {
    y_s <- (y - support[1]) / (support[2] - support[1])
    Z <- integrate(function(x) safe_exp(lambda[1]*x + lambda[2]*x^2 + lambda[3]*x^3), 0, 1)$value
    safe_exp(lambda[1]*y_s + lambda[2]*y_s^2 + lambda[3]*y_s^3) / (Z * (support[2] - support[1]))
  }
}

# Common yield grid (rescaled to [0, 1] to match normalized moments)
yield_grid <- seq(0, 1, length.out = 101)

# --- MBME Results from OLS moments ---
mbme_ols_sub <- list()

# Add regularization to prevent invalid moment combinations
grid[, mu2 := pmax(mu2, mu1^2 + 1e-4)]  # Ensures positive-definite second moment
grid[, mu3 := pmin(pmax(mu3, mu2 * mu1 + 1e-4), mu1)]  # Trim skewness for feasibility

for (i in seq_len(nrow(grid))) {
  mu1 <- grid$mu1[i]
  mu2 <- grid$mu2[i]
  mu3 <- grid$mu3[i]

  f_density <- tryCatch(
    estimate_mbme_density(mu1, mu2, mu3, support = c(0, 1)),
    error = function(e) {
      message("Failed at row ", i, ": ", conditionMessage(e))
      return(NULL)
    }
  )

  if (!is.null(f_density)) {
    dens_vals <- sapply(yield_grid, f_density)
    mbme_ols_sub[[i]] <- data.table(
      grid_row = i,
      yield = yield_grid * (yield_max - yield_min) + yield_min,  # Back-transform
      density = dens_vals / (yield_max - yield_min),  # Adjust for scale
      n = grid$n[i],
      prcp_t = grid$prcp_t[i],
      method = "Sub_OLS"
    )
  }
}

# --- MBME Results from GAM moments ---
mbme_gam_sub <- list()

grid[, mu2_gam := pmax(mu2_gam, mu1_gam^2 + 1e-4)]
grid[, mu3_gam := pmin(pmax(mu3_gam, mu2_gam * mu1_gam + 1e-4), mu1_gam)]

for (i in seq_len(nrow(grid))) {
  mu1 <- grid$mu1_gam[i]
  mu2 <- grid$mu2_gam[i]
  mu3 <- grid$mu3_gam[i]

  f_density <- tryCatch(
    estimate_mbme_density(mu1, mu2, mu3, support = c(0, 1)),
    error = function(e) {
      message("Failed at row ", i, ": ", conditionMessage(e))
      return(NULL)
    }
  )

  if (!is.null(f_density)) {
    dens_vals <- sapply(yield_grid, f_density)
    mbme_gam_sub[[i]] <- data.table(
      grid_row = i,
      yield = yield_grid * (yield_max - yield_min) + yield_min,
      density = dens_vals / (yield_max - yield_min),
      n = grid$n[i],
      prcp_t = grid$prcp_t[i],
      method = "Sub_GAM"
    )
  }
}

# Combine and label
mbme_results_ols_sub <- rbindlist(mbme_ols_sub, use.names = TRUE)
mbme_results_gam_sub <- rbindlist(mbme_gam_sub, use.names = TRUE)
mbme_results_sub <- rbindlist(list(mbme_results_ols_sub, mbme_results_gam_sub), use.names = TRUE)

# Label for plotting (revised to use only prcp_t as weather factor)
mbme_results_sub[, weather := paste0("prcp = ", round(prcp_t, 1))]
weather_levels <- mbme_results_sub[, sort(unique(weather))]
mbme_results_sub[, weather := factor(weather, levels = weather_levels)]
mbme_results_sub[, n_round := factor(round(n, 0), levels = sort(unique(round(n, 0))))]


saveRDS(mbme_results_sub, here("Data","Processed","mbme_results_sub.rds"))

```

## MBME estimation at field-level densities

```{r mbme-field-estimation}

# Reuse estimate_mbme_density() and yield_grid defined earlier

mbme_ols_field <- list()

for (i in seq_len(nrow(field_grid))) {
  mu1 <- field_grid$mu1[i]; mu2 <- field_grid$mu2[i]; mu3 <- field_grid$mu3[i]
  f_density <- tryCatch(
    estimate_mbme_density(mu1, mu2, mu3, support = c(0, 1)),
    error = function(e) NULL)
  if (!is.null(f_density)) {
    dens_vals <- sapply(yield_grid, f_density)
    mbme_ols_field[[i]] <- data.table(
      grid_row = i,
      yield = yield_grid * (yield_max - yield_min) + yield_min,
      density = dens_vals / (yield_max - yield_min),
      n = field_grid$n[i],
      prcp_t = field_grid$prcp_t[i],
      method = "Field-OLS"
    )
  }
}

# Estimate MBME densities from GAM-based field moments

mbme_gam_field <- list()

for (i in seq_len(nrow(field_grid))) {
  mu1 <- field_grid$mu1_gam[i]
  mu2 <- field_grid$mu2_gam[i]
  mu3 <- field_grid$mu3_gam[i]

  f_density <- tryCatch(
    estimate_mbme_density(mu1, mu2, mu3, support = c(0, 1)),
    error = function(e) NULL
  )

  if (!is.null(f_density)) {
    dens_vals <- sapply(yield_grid, f_density)
    mbme_gam_field[[i]] <- data.table(
      grid_row = i,
      yield = yield_grid * (yield_max - yield_min) + yield_min,
      density = dens_vals / (yield_max - yield_min),
      n = field_grid$n[i],
      prcp_t = field_grid$prcp_t[i],
      method = "Field-GAM"
    )
  }
}

# Combine both OLS and GAM field-level results

mbme_results_ols_field <- rbindlist(mbme_ols_field, use.names = TRUE)
mbme_results_gam_field <- rbindlist(mbme_gam_field, use.names = TRUE)
mbme_results_field <- rbindlist(list(mbme_results_ols_field, mbme_results_gam_field), use.names = TRUE)

# Label for plotting (revised to use only prcp_t as weather factor)
mbme_results_field[, weather := paste0("prcp = ", round(prcp_t, 1))]
weather_levels <- mbme_results_field[, sort(unique(weather))]
mbme_results_field[, weather := factor(weather, levels = weather_levels)]
mbme_results_field[, n_round := factor(round(n, 0), levels = sort(unique(round(n, 0))))]


saveRDS(mbme_results_field, here("Data","Processed","mbme_results_field.rds"))




```


## Combine all MBME results and save

```{r combine-mbme-results}

mbme_results_sub <- readRDS(here("Data","Processed","mbme_results_sub.rds"))
mbme_results_field <- readRDS(here("Data","Processed","mbme_results_field.rds"))
mbme_comb <- rbindlist(list(mbme_results_sub, mbme_results_field), use.names = TRUE, fill = TRUE)
mbme_comb[, n_round := factor(round(n, 0))]

saveRDS(mbme_comb, here("Data", "Processed", "mbme_results_combined.rds"))

```




## 4. Plot distribution of Yield-N x weather

```{r mbme-density-ols-vs-gam}

mbme_comb <- readRDS(here("Data","Processed","mbme_results_combined.rds"))

# Extract data level and method type
mbme_comb[, data_level := fifelse(grepl("^Sub", method), "Subplot", "Field")]
mbme_comb[, fit_method := fifelse(grepl("OLS", method), "OLS", "GAM")]

# Precipitation label
mbme_comb[, prcp_label := paste0("prcp = ", round(prcp_t, 1))]
mbme_comb[, prcp_label := factor(prcp_label, levels = sort(unique(prcp_label)))]

# Ensure consistent order
mbme_comb[, data_level := factor(data_level, levels = c("Subplot", "Field"))]
mbme_comb[, fit_method := factor(fit_method, levels = c("OLS", "GAM"))]

# Filter for 5 nitrogen levels
n_levels <- sort(unique(mbme_comb$n_round))
n_plot_levels <- n_levels[1:5]
mbme_plot <- mbme_comb[n_round %in% n_plot_levels]

# OLS only
mbme_ols <- mbme_plot[fit_method == "OLS"]

# Filter to one nitrogen rate 
target_n <- n_plot_levels[1]
mbme_nslice <- mbme_ols[n_round == target_n]

# Create discrete precipitation label
mbme_nslice[, prcp_label := paste0("prcp = ", round(prcp_t, 1))]
mbme_nslice[, prcp_label := factor(prcp_label, levels = sort(unique(prcp_label)))]

# Plot
yield_density_ols_n1 <- ggplot(
  mbme_nslice,
  aes(x = yield, y = density, color = prcp_label, group = interaction(prcp_label, data_level), linetype = data_level)
) +
  geom_line(size = 1.1, alpha = 0.95, lineend = "round") +
  scale_color_viridis_d(option = "D", name = "Precipitation Level") +
  scale_linetype_manual(values = c("Subplot" = "solid", "Field" = "dashed"), name = "Data Level") +
  labs(
    title = paste("OLS-Based Yield Distributions at N =", target_n),
    x = "Yield", y = "Density"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    strip.background = element_blank(),
    strip.text = element_text(size = 14)
  )

yield_density_ols_n1



ggsave( filename = here("Results","Figures", "ols_yield_density_n1.png"),
   plot = 
yield_density_ols_n1,
   width = 24,       # increase width (in inches)
   height = 16,       # adjust height as needed
   dpi = 300
 )


# GAM only
mbme_gam <- mbme_plot[fit_method == "GAM"]

# Filter to the same nitrogen rate as OLS
mbme_nslice_gam <- mbme_gam[n_round == target_n]

# Create discrete precipitation label
mbme_nslice_gam[, prcp_label := paste0("prcp = ", round(prcp_t, 1))]
mbme_nslice_gam[, prcp_label := factor(prcp_label, levels = sort(unique(prcp_label)))]

# Plot
yield_density_gam_n1 <- ggplot(
  mbme_nslice_gam,
  aes(x = yield, y = density, color = prcp_label, group = interaction(prcp_label, data_level), linetype = data_level)
) +
  geom_line(size = 1.1, alpha = 0.95, lineend = "round") +
  scale_color_viridis_d(option = "D", name = "Precipitation Level") +
  scale_linetype_manual(values = c("Subplot" = "solid", "Field" = "dashed"), name = "Data Level") +
  labs(
    title = paste("GAM-Based Yield Distributions at N =", target_n),
    x = "Yield", y = "Density"
  ) +
  theme_bw(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    strip.background = element_blank(),
    strip.text = element_text(size = 14)
  )

yield_density_gam_n1


ggsave(
  filename = here("Results", "Figures", "gam_yield_density_n1.png"),
  plot = yield_density_gam_n1,
  width = 24,
  height = 16,
  dpi = 300
)



```{r eonr estimation}

price_y <- 5.5  # price per unit yield (e.g., $/bushel)
price_n <- 0.7  # price per unit nitrogen (e.g., $/lb)

profit_dt <- mbme_comb[, {
  # Subset and get numeric integration over yield
  expected_profit <- sum(density * (price_y * yield - price_n * n)) * (yield[2] - yield[1])
  .(expected_profit = expected_profit)
}, by = .(n, prcp_t,data_level,fit_method)]


library(flextable)

#  How the estimated profit at a certain N x prcp_t will be different by data_level at a given fit_method?

# Wide format for OLS
ols_wide <- dcast(
  profit_dt[fit_method == "OLS"],
  n + prcp_t ~ data_level,
  value.var = "expected_profit"
)
ols_wide[, diff := Field - Subplot]
setnames(ols_wide, c("Subplot", "Field", "diff"), c("Subplot_Profit", "Field_Profit", "Field_Minus_Subplot"))

# Wide format for GAM
gam_wide <- dcast(
  profit_dt[fit_method == "GAM"],
  n + prcp_t ~ data_level,
  value.var = "expected_profit"
)
gam_wide[, diff := Field - Subplot]
setnames(gam_wide, c("Subplot", "Field", "diff"), c("Subplot_Profit", "Field_Profit", "Field_Minus_Subplot"))


# OLS Table
ft_ols <- flextable(ols_wide) %>%
  set_header_labels(
    n = "Nitrogen",
    prcp_t = "Precipitation",
    Subplot_Profit = "Subplot ($)",
    Field_Profit = "Field ($)",
    Field_Minus_Subplot = "Field - Subplot ($)"
  ) %>%
  colformat_num(j = c("Subplot_Profit", "Field_Profit", "Field_Minus_Subplot"), digits = 2) %>%
  autofit() %>%
  add_header_lines("Expected Profit by Data Level — OLS")

# GAM Table
ft_gam <- flextable(gam_wide) %>%
  set_header_labels(
    n = "Nitrogen",
    prcp_t = "Precipitation",
    Subplot_Profit = "Subplot ($)",
    Field_Profit = "Field ($)",
    Field_Minus_Subplot = "Field - Subplot ($)"
  ) %>%
  colformat_num(j = c("Subplot_Profit", "Field_Profit", "Field_Minus_Subplot"), digits = 2) %>%
  autofit() %>%
  add_header_lines("Expected Profit by Data Level — GAM")

library(webshot2)


# Save OLS table
save_as_image(ft_ols, path = "Results/Figures/expected_profit_OLS.png")

# Save GAM table
save_as_image(ft_gam, path = "Results/Figures/expected_profit_GAM.png")


```